---
title: "Exploratory Data Analysis"
author: "Trent Meyer"
date: "2022-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Use time series analysis
## Chapter 6 year*region interaction term
## Chapter 7: generalized linear model (glm())

```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(lubridate)
tick_adult <- read.csv(here("Data", "Deer_Tick_Surveillance__Adults__Oct_to_Dec__excluding_Powassan_virus__Beginning_2008.csv"))
```

Let's take a look at the data!
```{r}
head(tick_adult)
glimpse(tick_adult)
str(tick_adult)
```

Let's rename the parasite and bacteria variables, and also read in County and Year as factors instead of characters:
```{r}
tick_adult <- tick_adult %>% 
  rename(B.burgdorferi = "B..burgdorferi....",
         A.phagocytophilum = "A..phagocytophilum....",
         B.microti = "B..microti....",
         B.miyamotoi = "B..miyamotoi....") %>%
  mutate(County = as.factor(County),
         Year = as.factor(Year))
```

```{r}
# tick_adult <- tick_adult %>% mutate(County = as.factor(County), 
#                                     Year = paste0("01-01-", Year))

# tick_adult <- tick_adult %>% mutate(Year = mdy(Year),
#                                     Year = year(Year))
```

I think tick population density is a reasonable response variable. Let's graph tick population density to see what the distribution looks like.
```{r}
ggplot(data = tick_adult, aes(x = Tick.Population.Density)) +
  geom_histogram()
```
Lots of zeros!

We see that there are two that have a Tick.Population.Density > 750. I am not sure if this is a mistake, or if they actually found this high of a tick population density. I think I am going to remove these two observations.

Can you tell how they defined tick population density? Because if it is ticks/ha or something, then 750 might be (horribly) reasonabile. What are the units?

```{r}
tick_adult <- tick_adult %>% filter(Tick.Population.Density < 750)
```

Let's replot the distribution of Tick.Population.Density:
```{r}
ggplot(data = tick_adult, aes(x = Tick.Population.Density)) +
  geom_histogram()
```

We can see that the data is right-skewed.

```{r}
summary(tick_adult)
```

When I look at this, I see that the maximum is a bit higher than the mean. I think that I want to use this as my response for most of the statistical tests.

# Is there significant difference in Tick.Population.Density between the first year (2008), and last year (2021)?

I am going to subset the data to only have the observations for the years 2008 and 2021.

```{r}
tick_adult_small <- tick_adult %>% filter(Year == "2008" | Year == "2021")
```

Now I want to group by the year and find the mean Tick.Population.Density for both of the years.

```{r}
tick_adult_small_total <- tick_adult_small %>% group_by(Year) %>% summarise(mean_density = mean(Tick.Population.Density, na.rm = T))
```

We can see that 2021 is a bit higher than 2008! I now want to see if there is a significant difference between the two years.

```{r}
ggplot(data = tick_adult_small, aes(x = Year, y = Tick.Population.Density)) +
  geom_boxplot()

ggplot(data = tick_adult_small, aes(x = Tick.Population.Density)) +
  geom_histogram(bins = 10) +
  facet_wrap(~Year)
```

There is a bit of overlap in the spread of each year, however it looks like the mean is higher for 2021 compared to 2008. These plots are much easier to read with the few observations out that were making it difficult to read and compare between years. The histograms show that the peaks are in slightly different location on the x-axis.

I think the way I have set this up, I am going to have to run a two-sample t-test. 

Null Hypothesis: There is no true difference in mean Tick.Population.Density between 2008 and 2021.
Alternative Hypothesis: There is a true difference in mean Tick.Population.Density  between 2008 and 2021.

```{r}
year_ttest <- t.test(Tick.Population.Density ~ Year, data = tick_adult_small)
```

We can see from here that we have a p-value of `year_ttest$p.value`, which is very large! This would lead us to choose not to reject the null hypothesis, which means that we have evidence that there is no true difference in mean Tick.Population.Density between the years 2008 and 2021.

Since you have data on sites, I wonder if it would be worth doing this as a paired t-test where you have density in 2008 and 2021 as paired measures per site?

# Is there an association between Tick.Population.Density and Region (instead of county)?

```{r}
ggplot(data = tick_adult, aes(x = County, y = Total.Ticks.Collected)) +
  geom_col() +
  coord_flip()
```
St. Lawrence occurs twice. You need to check the levels for your factor variables to look for errors.

Also, total.ticks.collected is misleading, because if someone tried harder to collect ticks in e.g. Westchester, and gets more, it may not mean there are actually more ticks there. So it might be more meaningful to compare tick.population.density, which they have tried to make comparable (though they don't seem to tell us how in the metatdata)

This plot and question are difficult to answer so I think I am going to create a variable for Region, and test if there is a difference between the 9 regions in NYS. Also, we can see that there are two levels for St. Lawrence County, so I will combine those into the North Country region.

Good. And see what I wrote about using pop density rather than total ticks collected.

```{r}
tick_adult <- tick_adult %>% mutate(Region = case_when(
  County == "Suffolk" | County == "Nassau" ~ "Long Island",
  
  County == "Brooklyn" | County == "Bronx" | County == "Manhattan" | 
    County == "Staten Island" | County == "Queens" ~ "New York City",
  
  County == "Dutchess" | County == "Orange" | County == "Putnam" |
    County == "Rockland" | County == "Sullivan" | County == "Ulster" |
    County == "Westchester" ~ "Hudson Valley",
  
  County == "Niagara" | County == "Erie" | County == "Chautauqua" |
    County == "Cattaragus" | County == "Cattaraugus" | 
    County == "Allegany" ~ "Western New York",
  
  County == "Orleans" | County == "Genesee" | County == "Wyoming" |
    County == "Monroe" | County == "Livingston" | County == "Wayne" |
    County == "Ontario" | County == "Yates" | County == "Seneca" ~ "Finger Lakes",
  
  County == "Steuben" | County == "Schuyler" | County == "Chemung" | 
    County == "Tompkins" | County == "Tioga" | County == "Chenango" 
  | County == "Broome" | County == "Delaware" ~ "Southern Tier",
  
  County == "Cortland" | County == "Cayuga" | County == "Onondaga" | 
    County == "Oswego" | County == "Madison" ~ "Central New York",
  
  County == "St. Lawrence" | County == "St Lawrence" | County == "Lewis" | 
    County == "Jefferson" | County == "Hamilton" | County == "Essex" | 
    County == "Clinton" | County == "Franklin" ~ "North Country",
  
  County == "Oneida" | County == "Herkimer" | County == "Fulton" | 
    County == "Montgomery" | County == "Otsego" | 
    County == "Schoharie" ~ "Mohawk Valley",
  
  County == "Albany" | County == "Columbia" | County == "Greene" | 
    County == "Warren" | County == "Washington" | County == "Saratoga" | 
    County == "Schenectady" | County == "Rensselaer" ~ "Capital District"
))
```
Well done!


Oof, that was a lot about code, now let's replot this by Region!
```{r}
ggplot(data = tick_adult, aes(x = Region, y = Tick.Population.Density)) +
  geom_col() +
  coord_flip()
```

It looks like there are possible differences, with Capital Region being much higher, and also possibly Hudson Valley have a higher density? Long Island seems to have a pretty low tick density.

Since we have a categorical predictor with at least three levels (Region) and a quantitative response, I think the best test would be an ANOVA. First we have to fit the model:

Null Hypothesis: There is no difference in Region mean Tick.Population.Density
Alternative Hypothesis: There is a difference in Region mean Tick.Population.Density

```{r}
region_mod <- lm(Tick.Population.Density ~ Region, data = tick_adult)
```

Then we need to check the assumptions:

```{r}
autoplot(region_mod, smooth.colour = NA)
```

This is difficult because we do need see very much of a line in the Normal Q-Q plot, meaning the normality assumption is not perfect. Also, we do not see the constant variance in the Res vs Fitted plot, but this is a little bit better than the normality.

Smells like glm to me!


```{r}
region_mod_summary <- summary(region_mod)
region_anova <- anova(region_mod)
```

So from the ANOVA table, we can see a very small p-value of 2.656e-12, which means that there is evidence for a difference between the mean Tick.Population.Density by Region.

So we can do this as a glm and you can do tukey tests or similar to tell which regions are different.

# Is there a difference in the Tick.Population.Density between the years 2008 and 2021 and Region?

```{r}
tick_adult_total <- tick_adult %>% group_by(Region, Year) %>% summarise(total_ticks_collected = sum(Total.Ticks.Collected))
```

This gives us a little look into the possibility that there might be a difference in the number of ticks collected given the year and region.

```{r}
ggplot(data = tick_adult, aes(x = Region, y = Tick.Population.Density)) +
  geom_point() +
  coord_flip() +
  facet_wrap(~Year)
```

It looks like the Capital District has higher Tick.Population.Density for years 2009, 2011, and 2020. I might try to fit a linear model and predict the Tick.Population.Density collected given year and region. I want to add an interaction term between Year and Region.

```{r}
tick_adult_small <- tick_adult %>% mutate(Region = case_when(
  County == "Suffolk" | County == "Nassau" ~ "Long Island",
  
  County == "Brooklyn" | County == "Bronx" | County == "Manhattan" | 
    County == "Staten Island" | County == "Queens" ~ "New York City",
  
  County == "Dutchess" | County == "Orange" | County == "Putnam" |
    County == "Rockland" | County == "Sullivan" | County == "Ulster" |
    County == "Westchester" ~ "Hudson Valley",
  
  County == "Niagara" | County == "Erie" | County == "Chautauqua" |
    County == "Cattaragus" | County == "Cattaraugus" | 
    County == "Allegany" ~ "Western New York",
  
  County == "Orleans" | County == "Genesee" | County == "Wyoming" |
    County == "Monroe" | County == "Livingston" | County == "Wayne" |
    County == "Ontario" | County == "Yates" | County == "Seneca" ~ "Finger Lakes",
  
  County == "Steuben" | County == "Schuyler" | County == "Chemung" | 
    County == "Tompkins" | County == "Tioga" | County == "Chenango" 
  | County == "Broome" | County == "Delaware" ~ "Southern Tier",
  
  County == "Cortland" | County == "Cayuga" | County == "Onondaga" | 
    County == "Oswego" | County == "Madison" ~ "Central New York",
  
  County == "St. Lawrence" | County == "St Lawrence" | County == "Lewis" | 
    County == "Jefferson" | County == "Hamilton" | County == "Essex" | 
    County == "Clinton" | County == "Franklin" ~ "North Country",
  
  County == "Oneida" | County == "Herkimer" | County == "Fulton" | 
    County == "Montgomery" | County == "Otsego" | 
    County == "Schoharie" ~ "Mohawk Valley",
  
  County == "Albany" | County == "Columbia" | County == "Greene" | 
    County == "Warren" | County == "Washington" | County == "Saratoga" | 
    County == "Schenectady" | County == "Rensselaer" ~ "Capital District"
))

tick_adult_small <- tick_adult_small %>% filter(Year == "2008" | Year == "2021")
```


```{r}
full_mod <- glm(Tick.Population.Density ~ Region + Year + Region*Year, data = tick_adult_small)
summary(full_mod)
```

I can see here that the p-value for the Long Island Region is not significant at all, instead it is a very large p-value.
 
 
Can you think about how you might make a reasonable visualization of that test?

# Is there a difference between mean B.burgdorferi percentage in the years 2008 and 2021?
```{r}
ggplot(data = tick_adult_small, aes(x = B.burgdorferi)) +
  geom_histogram() +
  facet_wrap(~Year)
```

There looks to be a difference in where the peaks are on the y-axis, so this leads me to believe that there could be a significant difference in means between the years.

This will lead me to run a two-sample t-test with the following hypotheses:

Null Hypothesis: There is no difference in mean B.burgdorferi percentage between the years 2008 and 2021.
Alternative Hypothesis: There is a difference in mean B.burgdorferi percentage between the years 2008 and 2021.

```{r}
bburg_ttest <- t.test(B.burgdorferi ~ Year, data = tick_adult_small)
```

We can see a p-value of `bburg_ttest$p.value`, which is very small, and leads us to determine that we can reject our null, and see evidence that there is a significant difference between the mean B.burgdorferi percentage in the years 2008 and 2021.

# Is there a difference between mean A.phagocytophilum percentage in the years 2008 and 2021?
```{r}
ggplot(data = tick_adult_small, aes(x = A.phagocytophilum)) +
  geom_histogram() +
  facet_wrap(~Year)
```

Here it looks like the peak is higher for 2021, meaning the mean A.phagocytophilum percentage could possibly be significantly different compared to 2008. I will now test to see if there is a significant difference in the means.

This will lead me to run a two-sample t-test with the following hypotheses:

Null Hypothesis: There is no difference in mean A.phagocytophilum percentage between the years 2008 and 2021.
Alternative Hypothesis: There is a difference in mean A.phagocytophilum percentage between the years 2008 and 2021.

```{r}
aphago_ttest <- t.test(A.phagocytophilum ~ Year, data = tick_adult_small)
```

We can see a p-value of `aphago_ttest$p.value`, which is small, and leads us to determine that we can reject our null, and see evidence that there is a significant difference between the mean A.phagocytophilum percentage in the years 2008 and 2021.

# Is there an association between mean B.microti percentage in the years 2008 and 2021?
```{r}
ggplot(data = tick_adult_small, aes(x = B.microti)) +
  geom_histogram() +
  facet_wrap(~Year)
```

There is more spread for 2021, however the peaks both seem to be at or close to 0%. I think this one would show less significance than the other questions regarding bacteria.

This will lead me to run a two-sample t-test with the following hypotheses:

Null Hypothesis: There is no difference in mean B.microti percentage between the years 2008 and 2021.
Alternative Hypothesis: There is a difference in mean B.microti percentage between the years 2008 and 2021.

```{r}
bmicroti_ttest <- t.test(B.microti ~ Year, data = tick_adult_small)
```

We can see a p-value of `bmicroti_ttest$p.value`, which is small, and leads us to determine that we can reject our null, and see evidence that there is a significant difference between the mean B.microti percentage in the years 2008 and 2021.

# Is there an association between mean B.miyamotoi percentage in the years 2015 and 2021 (there was no data for 2008-2014)?
```{r}
tick_adult_small2 <- tick_adult %>% filter(Year == "2015" | Year == "2021")
ggplot(data = tick_adult_small2, aes(x = B.miyamotoi)) +
  geom_histogram() +
  facet_wrap(~Year)
```

This one is not as obvious that there could be a difference in the means. They both have peaks around zero, however there are some values in 2015 greater than 7.5%, which might make it have a higher mean, and cause the difference to be significance.

This will lead me to run a two-sample t-test with the following hypotheses:

Null Hypothesis: There is no difference in mean B.miyamotoi percentage between the years 2015 and 2021.
Alternative Hypothesis: There is a difference in mean B.miyamotoi percentage between the years 2015 and 2021.

```{r}
bmiya_ttest <- t.test(B.miyamotoi ~ Year, data = tick_adult_small2)
```

This is a very high p-value, meaning that there is not enough evidence to reject the null that there was no difference between the mean B.miyamotoi percentage between the years 2015 and 2021.

Running this many t-tests suggests to me that you may wish to use a Bonferroni correction. Here are some details: https://www.statology.org/bonferroni-correction/
