---
title: "Capdap Presentation"
author: "Trent Meyer"
date: "2022-12-02"
output: 
  revealjs::revealjs_presentation:
    theme: moon
    transition: none
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

## Introduction to my dataset

- My data comes from the [NYSDOH public database](https://health.data.ny.gov/Health/Deer-Tick-Surveillance-Adults-Oct-to-Dec-excluding/vzbp-i2d4)

- Active tick surveillance began statewide in 2008

- I chose this data because at home, I find lots of ticks on myself, my dogs, etc when in the woods

## Glimpse of the data set

```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(lubridate)
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
tick_adult <- read.csv(here("Data", "Deer_Tick_Surveillance__Adults__Oct_to_Dec__excluding_Powassan_virus__Beginning_2008.csv"))
head(tick_adult)
```

- Each row corresponds to a collection in a county
- Use Tick.Population.Density as response variable

## Is there an association between Tick.Population.Density and Year?

```{r}
tick_adult <- tick_adult %>% mutate(Region = case_when(
  County == "Suffolk" | County == "Nassau" ~ "Long Island",
  
  County == "Brooklyn" | County == "Bronx" | County == "Manhattan" | 
    County == "Staten Island" | County == "Queens" ~ "New York City",
  
  County == "Dutchess" | County == "Orange" | County == "Putnam" |
    County == "Rockland" | County == "Sullivan" | County == "Ulster" |
    County == "Westchester" ~ "Hudson Valley",
  
  County == "Niagara" | County == "Erie" | County == "Chautauqua" |
    County == "Cattaragus" | County == "Cattaraugus" | 
    County == "Allegany" ~ "Western New York",
  
  County == "Orleans" | County == "Genesee" | County == "Wyoming" |
    County == "Monroe" | County == "Livingston" | County == "Wayne" |
    County == "Ontario" | County == "Yates" | County == "Seneca" ~ "Finger Lakes",
  
  County == "Steuben" | County == "Schuyler" | County == "Chemung" | 
    County == "Tompkins" | County == "Tioga" | County == "Chenango" 
  | County == "Broome" | County == "Delaware" ~ "Southern Tier",
  
  County == "Cortland" | County == "Cayuga" | County == "Onondaga" | 
    County == "Oswego" | County == "Madison" ~ "Central New York",
  
  County == "St. Lawrence" | County == "St Lawrence" | County == "Lewis" | 
    County == "Jefferson" | County == "Hamilton" | County == "Essex" | 
    County == "Clinton" | County == "Franklin" ~ "North Country",
  
  County == "Oneida" | County == "Herkimer" | County == "Fulton" | 
    County == "Montgomery" | County == "Otsego" | 
    County == "Schoharie" ~ "Mohawk Valley",
  
  County == "Albany" | County == "Columbia" | County == "Greene" | 
    County == "Warren" | County == "Washington" | County == "Saratoga" | 
    County == "Schenectady" | County == "Rensselaer" ~ "Capital District"
))

(ggplot(data = tick_adult, aes(x = Year, y = Tick.Population.Density, group = Region)) +
  geom_point())
```

## Narrowing it down to 2008 and 2021

```{r}
tick_adult_small <- tick_adult %>% mutate(Region = case_when(
  County == "Suffolk" | County == "Nassau" ~ "Long Island",
  
  County == "Brooklyn" | County == "Bronx" | County == "Manhattan" | 
    County == "Staten Island" | County == "Queens" ~ "New York City",
  
  County == "Dutchess" | County == "Orange" | County == "Putnam" |
    County == "Rockland" | County == "Sullivan" | County == "Ulster" |
    County == "Westchester" ~ "Hudson Valley",
  
  County == "Niagara" | County == "Erie" | County == "Chautauqua" |
    County == "Cattaragus" | County == "Cattaraugus" | 
    County == "Allegany" ~ "Western New York",
  
  County == "Orleans" | County == "Genesee" | County == "Wyoming" |
    County == "Monroe" | County == "Livingston" | County == "Wayne" |
    County == "Ontario" | County == "Yates" | County == "Seneca" ~ "Finger Lakes",
  
  County == "Steuben" | County == "Schuyler" | County == "Chemung" | 
    County == "Tompkins" | County == "Tioga" | County == "Chenango" 
  | County == "Broome" | County == "Delaware" ~ "Southern Tier",
  
  County == "Cortland" | County == "Cayuga" | County == "Onondaga" | 
    County == "Oswego" | County == "Madison" ~ "Central New York",
  
  County == "St. Lawrence" | County == "St Lawrence" | County == "Lewis" | 
    County == "Jefferson" | County == "Hamilton" | County == "Essex" | 
    County == "Clinton" | County == "Franklin" ~ "North Country",
  
  County == "Oneida" | County == "Herkimer" | County == "Fulton" | 
    County == "Montgomery" | County == "Otsego" | 
    County == "Schoharie" ~ "Mohawk Valley",
  
  County == "Albany" | County == "Columbia" | County == "Greene" | 
    County == "Warren" | County == "Washington" | County == "Saratoga" | 
    County == "Schenectady" | County == "Rensselaer" ~ "Capital District"
))
tick_adult_small <- tick_adult_small %>% filter(Year == "2008" | Year == "2021")
(ggplot(data = tick_adult_small, aes(x = Year, y = Tick.Population.Density, group = Region)) +
  geom_point())
```

## Fitting my model
```{r}
full_mod <- glm(Tick.Population.Density ~ Region + Year + Region*Year, data = tick_adult_small)
(summary(full_mod))
```

## Final Plot
```{r}
tick_adult_small_mean <- tick_adult_small %>% group_by(Year) %>% summarise(mean = mean(Tick.Population.Density, na.rm=T))
(ggplot(data = tick_adult_small, aes(x = Year, y = Tick.Population.Density)) +
  geom_point() +
    geom_point(data = tick_adult_small_mean, aes(x = Year, y = mean), colour = "red", size = 3)
  )
```


## Is there a difference between mean B.burgdorferi percentage in the years 2008 and 2021?

```{r}
ggplot(data = tick_adult_small, aes(x = B..burgdorferi....)) +
  geom_histogram() +
  facet_wrap(~Year)
```

## Two sample t-test

```{r}
(bburg_ttest <- t.test(B..burgdorferi.... ~ Year, data = tick_adult_small))
```

## Final Plot
```{r}
tick_adult_bburg <- tick_adult_small %>% filter(Year == "2008" | Year == "2021")
tick_adult_bburg_mean <- tick_adult_bburg %>% group_by(Year) %>% summarise(mean = mean(B..burgdorferi...., na.rm=T))
ggplot(data = tick_adult_small, aes(x = Year, y = B..burgdorferi....)) +
  geom_point() + 
  geom_point(data = tick_adult_bburg_mean, aes(x = Year, y = mean), colour = "red", size = 3) 
```


## Conclusion

- Relearn many of these tests and analyses
- Which models and tests one should use are different for everyone
- Find helpful ways to plot after we create our model